Quasi-universal acceptance of basic science in the US

### 2025-07-07 preprint.Rmd
second round revisions

  Substantial minorities of the population report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g. the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1%), but also in every sub-sample (e.g., no trust in science: 87.3%; complete endorsement of flat Earth theory: 87.2%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
keywords          : Trust in science; conspiracy theories; conspiracy thinking; science knowledge.
floatsintext      : yes
linenumbers       : no 
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "doc" # "doc" for nice look, "man" for manuscripty
output: 
  papaja::apa6_pdf: 
    keep_tex: true
  papaja::apa6_docx: default
always_allow_html: true
appendix:
bibliography: references.bib
---

@@

Unfortunately, people who report a high degree of trust in science are a minority in most countries, and they are outnumbered by people who have low trust in science in many areas, e.g., most of Africa and significant parts of Asia [@wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020]. Moreover, trust in science has recently been declining in some countries [@alganTrustScientistsTimes2021; although see @wellcomeglobalmonitorPublicTrustScientists2021], including the US [@brianAmericansTrustScientists2023], where it is also increasingly polarizing [@gauchatPoliticizationSciencePublic2012; @krauseTrendsAmericansTrust2019a; @liPolarizationPublicTrust2022]. Rejection of science can take the more extreme form of conspiracy theories, many of which question the scientific consensus on vaccination, climate change, and even the shape of the Earth.

What does this apparent rejection of science actually entail? Do people who say they do not trust science, or who believe in conspiracy theories at odds with well-established science, reject most of science? Or, on the contrary, do they object to a few specific facets of science, while still accepting the overwhelming majority of basic science? Answering this question has theoretical implications: if some people reject science wholesale, then it’s possible that their distrust of science causes their rejection of specific scientific theories, or facilitates their acceptance of conspiracy theories; by contrast, if everyone trusts most of science, then the mistrust of specific facets of science is more likely to be post-hoc, a rationalization of other beliefs or behaviors. The present question also has practical implications: many communication attempts leverage the scientific consensus [e.g. on vaccination, climate change, etc., for review, see @vanstekelenburgScientificConsensusCommunicationContested2022; see also @veckalov27countryTestCommunicating2024]. These attempts are more likely to be successful if everyone trusts basic science than if some people reject science wholesale.

Studies of science knowledge provide relevant evidence, as people who answer science knowledge questions correctly presumably trust scientists in this respect. The average level of science knowledge is quite low, even in countries with extensive science education. For example, only 55% in the US (in 2014) and only 46% in the EU (in 2005) knew that antibiotics kill only bacteria and not viruses [@nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016]. However, these results only provide a lower bound on how much people trust basic science, as participants who do not provide the correct answer might accept it when it is pointed out to them.

@@

In a series of four pre-registered online studies (total n = 782), we asked US participants questions about well-established scientific facts. For each question, we asked participants what they thought the correct answer was (testing their knowledge of science), we informed them of the scientifically accepted answer, and asked them whether they accepted it (measuring their trust in basic science). We also measured participants’ trust in science using standard measures, as well as their beliefs in various conspiracy theories and their tendency to engage in conspiratorial thinking. The four studies, including materials, hypotheses, and analyses, were pre-registered and all materials and data are accessible via the Open Science Framework (https://osf.io/8utsj/?view_only=364478f01fc548d0a42b9962918d871d). The differences between the four studies are summarized presently, and the methods are detailed below.

@@

US participants were asked whether they accepted scientifically consensual answers on basic science questions.We found quasi-universal acceptance of basic science, with an overall rate of acceptance of `r combined_descriptives$mean_acceptance$Yes$share`, which remained very high for participants who declared not trusting science at all (`r combined_descriptives$acceptance_by_trust$'1'$Yes$share` of acceptance), or who endorsed theories blatantly violating scientific knowledge, such as flat earth (`r combined_descriptives$acceptance_by_conspiracy_belief$Yes$flatearth$'9'$share` of acceptance).

Overall, the present findings support the motivated rejection of science model [@lewandowskyMotivatedRejectionScience2016], in which “people tend to reject findings that threaten their core beliefs or worldview” (p. 217). A number of participants in our studies rejected basic tenets of science (on evolution, the shape of the earth, etc.). However, they still accepted the vast majority of basic scientific knowledge presented to them. This suggests that these participants had reasons to reject specific scientific knowledge, and that this rejection prompted them to express a lower trust in science when asked general questions on the topic. These participants did not appear to have general grounds for distrusting science, which should have led them to reject most or all of the science knowledge presented to them. These results are in line with several recent studies in which conspiracy theorists are not less susceptible to social influence in general [@pummererConspiracyBeliefsMajority; @altayConspiracyBelieversClaim2023].

Our findings are also relevant for the knowledge–attitudes model of trust in science [see, e.g., @bauerWhatCanWe2007]. The fact that even people with low science knowledge trust most of basic science might be taken as showing the power of even a modicum of science knowledge; on the other hand, it also means that attempts at increasing science knowledge in general might not have much effect on trust in science, by contrast with targeting the specific beliefs people are motivated to reject.

Finally, some of the present results also speak to the alienation model [@gauchatCulturalAuthorityScience2011], and more specifically to the need for epistemic autonomy [@frickerTestimonyEpistemicAutonomy2006]. Declaring not trusting science, or endorsing conspiracy theories [@harrisConspiracyTheoriesPopulism2023] might reflect a desire to maintain epistemic autonomy and not appear to ‘blindly’ accept epistemic authority. The acceptance of basic science facts would be reconciled with this need to appear epistemically autonomous by the suggestion that the acceptance stems from independent evaluation instead of trust (however implausible that might be: it’s not clear how participants could independently verify the ratio of glial cells to neurons, say, and few participants appeared to have engaged in even simple forms of verification).

In applied terms, the present results suggest the existence of a vast reservoir of trust in basic science that can be tapped into in science communication. Since people appear so resistant to rejecting basic science, stressing the basic science components of publicly controversial fields, from GMOs to climate change, might help reduce the ‘consensus gaps’ observed in these domains [on climate change, see, e.g. @ranneyClimateChangeConceptual2016].

The present studies have a number of limitations, in particular the lack of representative samples, and the focus on a single country. We hope that future studies will extend our findings to other populations.

@@

>>>>

  Many people report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science in substantial segments of the population. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g. the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1%), but also in every sub-sample (e.g., no trust in science: 87.3%; complete endorsement of flat Earth theory: 87.2%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
keywords          : Trust in science; conspiracy theories; conspiracy thinking; science knowledge.
floatsintext      : yes
linenumbers       : no 
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "doc" # "doc" for nice look, "man" for manuscripty
output: 
  papaja::apa6_pdf: 
    keep_tex: true
  papaja::apa6_docx: default
always_allow_html: true
appendix:
  - "appendix_acceptance_knowledge.Rmd"
  - "appendix_conspiracy_thinking.Rmd"
  - "appendix_trust.Rmd"
  - "appendix_conspiracy_belief.Rmd"
  - "appendix_knowledge.Rmd"
  - "appendix_by-question.Rmd"
  - "appendix_conspiracy-scales.Rmd"
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"
bibliography: references.bib
---

@@

Although recent global evidence shows that trust in science is moderately high [@colognaTrustScientistsTheir2024], it is far from being at ceiling. Large-scale polls have shown that people who report a high degree of trust in science are a minority in most countries, and they are outnumbered by people who have low trust in science in many areas, e.g., most of Africa and significant parts of Asia [@wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020]. Moreover, trust in science has recently been declining in some countries [@alganTrustScientistsTimes2021; @brianAmericansTrustScientists2023; although see @wellcomeglobalmonitorPublicTrustScientists2021; and @funkPublicConfidenceScientists2020] and in the US it is increasingly polarizing [@gauchatPoliticizationSciencePublic2012; @krauseTrendsAmericansTrust2019a; @liPolarizationPublicTrust2022].

Besides low answers on general trust in science questions, another indicator of distrust in science is the belief in conspiracy theories that question the scientific consensus on issues such as vaccination, climate change, and even the shape of the Earth. Conspiracy theories–in the realm of science or elsewhere–typically accuse a small group of powerful people to pursue nefarious goals in secrecy [@douglasUnderstandingConspiracyTheories2019; @medeSciencerelatedPopulismConceptualizing2020]. Some of these conspiracy theories are widespread [@rutjensConspiracyBeliefsScience2022]. In 2023, a survey in eight different countries found that up to 24% of respondents agreed that “climate change is a hoax and scientists touting its existence are lying” [@stockemerUnderstandingClimateChange2024]. Similarly, in 2021, 40% of Americans believed that “the dangers of genetically-modified foods are being hidden from the public” [down from 45% in 2020, @uscinskiHaveBeliefsConspiracy2022]. People who hold such views not only reject the relevant scientifically consensual facts, but also tend to believe in other conspiracy theories [@lewandowskyRoleConspiracistIdeation2013; @hornseyPsychologicalRootsAntivaccination2018; @hornseyRelationshipsConspiratorialBeliefs2018], and tend to say that they distrust science more generally [@vranicDidMyOwn2022; @stockemerUnderstandingClimateChange2024]. In spite of these correlations, the causal relationship between declaring general distrust in science and believing in one or several anti-science conspiracy theories is not clear. Although conspiracy thinking has been identified as a "root cause" of anti-science attitudes [@hornseyAttitudeRootsJiu2017] in the past, these claims rest largely on observational data.

What does this apparent lack of trust in science actually entail? Do people who say they do not trust science, or who believe in conspiracy theories at odds with well-established science, reject most of science? Or, on the contrary, do they object to a few specific facets of science, while still accepting the overwhelming majority of basic science?

A common conception of trust is a willingness to be vulnerable to another party, whether an individual, a group, or an institution [@mayerIntegrativeModelOrganizational1995a; @rousseauIntroductionSpecialTopic1998]. Accordingly, trust in science has been defined as “one’s willingness to rely on science and scientists (as representatives of the system) despite having a bounded understanding of science” [@wintterlinPredictingPublicTrust2022, p. 2]. Past research has disentangled this general concept of trust in science in various ways. Some research has identified different components of trust, the number of which varies, but which generally cover an epistemological and ethical dimension [@wilholtEpistemicTrustScience2013; @intemannScienceCommunicationPublic2023]. For example, @hendriksMeasuringLaypeoplesTrust2015 suggest distinguishing between expertise/competence, integrity, and benevolence, while @besleyReassessingVariablesUsed2021 add openness. Other research has highlighted differences in trust between scientific disciplines [@altenmullerExplainingPolarizedTrust2024; @gligoricHowSocialEvaluations2024; @gauchatCulturalCognitiveMappingScientific2018]. However, little research has assessed trust in specific scientific findings, besides contentious topics such as vaccines [e.g. @hornseyPsychologicalRootsAntivaccination2018], climate change [e.g. @stockemerUnderstandingClimateChange2024], evolution [e.g. @nadelsonTrustScienceScientists2015], genetically modified organisms (GMOs) [e.g. @fernbachExtremeOpponentsGenetically2019], or a combination of such topics [see e.g. @lewandowskyRoleConspiracistIdeation2013]. To the best of our knowledge, no research has investigated the extent to which people trust basic science facts (e.g. electrons are smaller than atoms). An extensive literature on science literacy has assessed whether people know such facts [@nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016], but not whether they accept the facts once presented to them.

Why does it matter whether trust in science is or is not related to trust in specific scientific findings? First, this question has theoretical implications. According to the most prominent explanation of (dis)trust in science–the deficit model–science knowledge is the main driver of attitudes towards science in general. A prediction of this model is that average trust in science on specific facts should be strongly associated with general trust in science. By contrast, a disconnect between the two would be in line with motivated reasoning accounts of trust in science. According to these accounts, science rejection serves to maintain coherence with other beliefs or behaviors [@lewandowskyMotivatedRejectionScience2016; @hornseyWhyFactsAre2020]. For example, someone might say they do not trust science in general because they are skeptical towards vaccines, not because they actually distrust most of science. A prediction of the motivated reasoning account is that general trust in science should be strongly correlated with conspiracy beliefs. Second, the present question has practical implications: Many communication attempts leverage the scientific consensus [e.g. on vaccination, climate change, etc., for review, see @vanstekelenburgScientificConsensusCommunicationContested2022; see also @veckalov27countryTestCommunicating2024]. These attempts are more likely to be successful if everyone trusts basic science than if some people reject science wholesale.

@@

In a series of four pre-registered online studies (total n = 782), we asked US participants questions about well-established, consensual scientific facts. For each question, we asked participants what they thought the correct answer was (testing their knowledge of science), we informed them of the scientifically accepted answer, and asked them whether they accepted it (measuring their trust in basic science). We also measured participants’ trust in science using standard measures, as well as their beliefs in various conspiracy theories and their tendency to engage in conspiratorial thinking. The four studies, including materials, hypotheses, and analyses, were pre-registered and all materials and data are accessible via the Open Science Framework (<https://osf.io/8utsj/>). The differences between the four studies are summarized presently, and the methods are detailed below.

@@

In four studies, we asked US participants whether they accepted scientifically consensual answers on basic science questions. We found quasi-universal acceptance of basic science, with an overall rate of acceptance of `r combined_descriptives$mean_acceptance$Yes$share`, which remained very high for participants who declared not trusting science at all (`r combined_descriptives$acceptance_by_trust$'1'$Yes$share` of acceptance), or who endorsed theories blatantly violating scientific knowledge, such as flat earth (`r combined_descriptives$acceptance_by_conspiracy_belief$Yes$flatearth$'9'$share` of acceptance).

This disconnect between declared general trust in science and average trust in (basic) scientific findings goes against predictions from the knowledge–attitudes model of trust in science [see, e.g., @bauerWhatCanWe2007], which posits that knowledge of scientific results is the main cause of trust in science. In line with past research [@allumScienceKnowledgeAttitudes2008], we find that both knowledge and acceptance of science are only weakly correlated with general trust in science. This supports recent definitions of science literacy, which suggest that for science literacy to be a meaningful concept for science attitudes, it needs to go beyond measuring knowledge of isolated science facts [@nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016].

By contrast, our findings align with a motivated reasoning account of trust in science [@lewandowskyMotivatedRejectionScience2016], in which “people tend to reject findings that threaten their core beliefs or worldview” (p. 217). A number of participants in our studies endorsed specific conspiracy theories questioning basic tenets of science (on evolution, the shape of the earth, etc.), while still accepting the vast majority of basic scientific facts presented to them. This suggests that these participants had reasons to reject only specific scientific knowledge, and that this rejection prompted them to express a lower trust in science when asked general questions on the topic. Consistent with this explanation, we found a strong association between belief in anti-science conspiracy theories and general trust in science.

Some of the present results also speak to the alienation model [@gauchatCulturalAuthorityScience2011], and more specifically to the need for epistemic autonomy [@frickerTestimonyEpistemicAutonomy2006]. Declaring not trusting science, or endorsing conspiracy theories [@harrisConspiracyTheoriesPopulism2023] might reflect a desire to maintain epistemic autonomy and not appear to ‘blindly’ accept epistemic authority. Such a need to appear epistemically autonomous could be reconciled with the acceptance of basic science facts if it was thought to stem from independent evaluation instead of trust. The majority of participants in Studies 3 and 4, two samples of vaccine-skeptical participants who scored high across a range of different conspiracy beliefs (see ESM), tended to claim that they had accepted the scientific consensus based on their own evaluation, however implausible that might be: it is not clear how participants could independently verify, say, the ratio of glial cells to neurons. These participants did also not engage (more than others) in simple forms of verification, i.e. clicking on sources provided in the answers. This result aligns with other results on conspiracy thinking: Recent studies have shown that, although they claim not to be, conspiracy theorists are in fact just as susceptible to social influence as others [@pummererConspiracyBeliefsMajority2024; @altayConspiracyBelieversClaim2023].

In applied terms, the present results have implications for science communication. Across various domains of science knowledge such as climate change [@veckalov27countryTestCommunicating2024] or vaccination [@salmonVaccineHesitancyCauses2015], researchers have observed a consensus gap: a gap between the scientific consensus and public opinion. Some of the most worrying consensus gaps relate to climate change, as substantial segments of the population disagree with scientists on what is happening and what to do about it [e.g., @eganClimateChangeUS2017]. Yet, we found that for much of basic, non-contentious science, such consensus gaps are absent: The vast majority of our participants--including two vaccine-skeptical samples (Studies 3 and 4)--did not appear to have general grounds for distrusting science, which should have led them to reject most or all of the science knowledge presented to them. Since even people who say they distrust science, or who reject specific scientific facts appear to accept most of basic science, stressing the basic science components of publicly controversial fields, from GMOs to climate change, might help reduce the consensus gaps observed in these domains [on climate change, see, e.g. @ranneyClimateChangeConceptual2016].

The present studies have a number of limitations, in particular the lack of representative samples, and the focus on a single country. While our results support motivated reasoning accounts of trust in science, they leave important questions unaddressed, in particular: Why do people reject specific scientific findings but not others? Suggestions have already been made for a number of issues such as vaccination [@mitonCognitiveObstaclesProVaccination2015], GMOs [@blanckeFatalAttractionIntuitive2015], or nuclear energy [@hacquinDisgustSensitivityPublic2021]. However, research is still needed to better understand what motivates these rejections [see e.g., @hornseyWhyFactsAre2020]. Moreover, rejection of scientific facts can manifest in various ways, for example calling into question the integrity of scientists [@medeSciencerelatedPopulismConceptualization2023a; @medeWhoSupportsSciencerelated2022], or denying the very possibility of scientifically investigating certain issues [@munroScientificImpotenceExcuse2010]. The classification of justifications we present here does not address these processes in detail. Finally, by design, most of the basic science knowledge presented in this study didn't directly relate to anything controversial for most people. Future studies could look at basic science that does relate to scientific topics which are the object of public and policy-relevant debates (e.g. GM food, climate science, vaccines), or scientific findings for which there is a more obvious potential conflict of interest (e.g. non-publicly funded medical research).

### Data availability

Data for all experiments and the simulations is available on the OSF project page (https://osf.io/8utsj/).

### Code availability

The code used to create all results (including tables and figures) of this manuscript is also available on the OSF project page (https://osf.io/8utsj/).

### Competing interest

The authors declare having no competing interests.

@@

```{r child = "appendix_conspiracy-scales.Rmd"}
```
